---
title: "Data Visualisation  \n &nbsp;  \nAssignment 4   \n &nbsp;  \nVisualising the PISA reading scores"
author: "Rahul Kumar (20230113)"
date: "MSc. Data Analytics"
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{center}
    \includegraphics[width=2in,height=2in]{logo.jpg}\LARGE\\}
  - \posttitle{\end{center}}
output: 
  pdf_document:
    toc: true
    number_sections: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

\pagebreak

<!-- # Introduction -->
<!-- The report visualizes the **PISA** (Programme for International Student Assessment) **reading performance** of 15-year old students, published by **OECD** (Organization for Economic Co-operation and Development) among it's various member states. The dataset is present in the file *OECD_PISA.csv* with the following columns: -->

<!-- - LOCATION : Names of the *member Country* -->
<!-- - SUBJECT  : The gender of the students, *BOY*, *GIRL*, and *TOT* which provides average of gender. -->
<!-- - TIME     : The *year* when the data was recorded. -->
<!-- - Value    : The *reading performance score* as calculated by the organization. -->

# PART 1 : Summary of the visualisation decisions 

The Plot provided in the assignment compares the *PISA reading performance for 2018 between boys and girls*  for the OECD member states. Below is the detailed explanation for the various aspects of the plot.

## Aesthetic Mappings

- The *Boys* and *Girls* data points are mapped to different shapes, allowing us to easily distinguish the genders.
- The *Boys* and *Girls* *shapes* are then colored using *3 different colors* - The *Average values* for the year 2018 are colored as *black*, The values for *Ireland* are colored as *red*, and rest of the data points are colored with a *dark gray type* color. This is done to attract the reader's attention to the *average* and *Ireland* values and make them stand out for easier comparison with other countries. 
- The *Boys* and *Girls* data points for a country is connected with a *line segment*. This is done to aid the reader in *visualizing the difference* in score among the gender of a country and compare them easily with other countries.

## Axes

- The x-axis holds *discrete values - country names* and is arranged in *ascending order of Boys reading score*. This is done to communicate a sense of order and keep the plot organized.
- The x-axis labels are *rotated at an angle* to facilitate easy reading of the country names. The labels text are then *colored same as their data points* to maintain consistency in the plot.
- The y-axis has a *continuous scale* ranging from *340-560 with an interval of 20 points* and is plotted inside the graph.
- The Tick marks for x and y-axis are removed.

## Gridlines

- The minor gridlines for x and y axis are removed to avoid cluttering.
- The major gridlines for x-axis is removed and y-axis is made thin to have a more cleaner plot.
- A *vertical line segment* is added starting from y=340 going till the datapoint of *Boys* for each country. This acts as a pseudo major gridline at x-axis.  

## Legend

- The legend is shifted to the bottom-left side of the graph.
- The legend title is removed and the background is colored in white.
- Only the general colored *Boys* and *Girls* are shown and not the average or Ireland colored datapoints. 

## Other Visual Details

- The plot background is light blue color.
- The Title and subtitle are written in same line, with font size of title being larger than subtitle.
- There is a blue line segment plotted at the top of the plot under the title and subtitle.

# PART 2 : Reproducing the Plot

In this section of the report I try to reproduce the plot provided in the assignment. Below is the detailed explanation on each aspect of the plot.

## Explanation

The *OECD_PISA.csv* contains data collected from 2000 - 2018. This plot requires us to use the data-points for *Boys* and *Girls* for the year 2018. Below are the data wrangling steps performed to get the required data.

1. The original country names are extracted by using "countrycode" library, and is used to replace the abbreviated LOCATION present in the dataset.
2. Some of the LOCATION do not match with the country names in the plot. Used *case_when* to change the LOCATION names to reflect the plot accordingly.
3. Changed the SUBJECT names from *BOY* to *Boys* and *GIRL* to *Girls* to reflect the plot.
4. Filter the data for the *year = 2018* where the SUBJECTS are *Boys and Girls*.
5. Added another 2 columns to reflect the *Boys* and *Girls* score for each country - This is used when adding a *line segment* between the genders and between the base y=340 to each of the boys value per country.
6. Arranged the filtered dataset in ascending order of the *Boys* column. - This is done to reflect the ordered x-axis.
7. Added 1 more column which reflects the *color* of x-axis text, i.e. when *LOCATION = 'OECD - Average' use black* or if *LOCATION = 'Ireland' use red*, else use "#7F7F7F".
8. changed the SUBJECT value for *Ireland and OECD - Average* in-order to have *6 different SUBJECT factors* - namely: *Boys, Boys-avg, Boys-IRL, Girls, Girls-avg, and Girls-IRL*. This is done so that we can color each factor differently as required by the plot.
9. Finally convert the LOCATION and SUBJECT column to factors as they are discrete variables which are currently represented as vector variables. 

### Aesthetic Mappings

- Used geom_point to map *x-axis with the reordered LOCATION*, *y-axis with the Values(scores)*, and the *SUBJECT* is mapped to *shape, color and fill*.
- *scaled_shape_manual* to provide the target shape for the SUBJECTS as : **19** for *all 3 Boys* factors and **23** for *all 3 girls* factors.
- *scaled_color_manual* to provide the target color as: **Boys = "#406D89", Boys-avg = "black", Boys-IRL = "red"** and similarly for girls, girls-avg and girls-IRL.
- Used *scaled_fill_manual* to color the inside of the data-point. used : **Boys = "#E2EDF3", Boys-avg = "black", Boys-IRL = "red"** and similarly for girls, girls-avg and girls-IRL.
- adding *line_segment* between each data points of the country's Boys and Girls using *geom_segment* and coloured it as "#C4D2DB". 

### Axes

- The x-axis labels are *rotated at an angle of 45 degree*. 
- The labels text are then colored same as their data points by using  *PISA_2018$text_Colour[c(T,F)]*. This filters the dataset on text-color column and recycles the value to only pick 1 color for the country value.
- The y-axis has a *continuous scale* ranging from *340-560 with an interval of 20 points* and is plotted inside the graph by *expanding* the x-axis using *expand(0,2.2)*. This accommodate extra space and avoids overlapping of y axis labels and data-points.
- The Tick marks for x and y-axis are removed.

### Gridlines

- The minor gridlines for x and y axis are removed to avoid cluttering and the major gridlines for x-axis is removed and y-axis is made thin to have a more cleaner plot.
- A *vertical line segment* is added starting from y=340 going till the datapoint of *Boys* for each country in white colour. This acts as a pseudo major gridline at x-axis.

### Legend

- The legend title is removed and background is made as white and moved to bottom left corner of the plot using the legend.* function in theme of ggplot.
- used breaks = c("Boys", "Girls") in all manual_scales to override the legend keys.

### Other visual details

- The background of the plot is "E2EDF3" - a light blue color.
- The plot title and subtitle uses font : "Times" and are adjusted using hjust and vjust and margin to make it appear in 1 single line.
- Added a blue line segment at the top of the plot using *annotate* function.
- **Was not able to recreate the exact plot title and subtitle. Used the font which was closest to the image.**

## Plot

```{r code_P2}
library(tidyverse)
library(countrycode)

# read the csv file
PISA <- read.csv("./OECD_PISA.csv", fileEncoding = "UTF-8-BOM") %>% select(c(1,3,6,7))

# use countrycode library to change the LOCATION names to full name
PISA$LOCATION <- countrycode(PISA$LOCATION, "iso3c", "country.name")

# mutate the dataset to change the country names and subject levels to match the plot
PISA <- PISA %>% mutate(LOCATION = case_when( is.na(LOCATION) ~ 'OECD - Average',
                                     LOCATION == 'Slovakia' ~ 'Slovak Republic',
                                     LOCATION == 'Czechia'  ~ 'Czech Republic',
                                     LOCATION == 'South Korea' ~ 'Korea',
                                     TRUE ~ LOCATION),
                        SUBJECT = case_when( SUBJECT == 'BOY' ~ 'Boys',
                                     SUBJECT == 'GIRL' ~ 'Girls',
                                     TRUE ~ SUBJECT))

# filter 2018 data and add 2 columns Boys and Girls - useful for plotting line
# segment between the genders for each country
PISA_2018 <- PISA %>% filter(TIME == 2018, SUBJECT != 'TOT') %>% 
             group_by(LOCATION) %>% mutate( Boys = Value[SUBJECT == 'Boys'], 
                                            Girls = Value[SUBJECT == 'Girls']) %>%
             arrange(Boys) %>%  ungroup()

# add text_colour as a column -- used for assigning x-axis label text color
PISA_2018 <- mutate(PISA_2018, text_Colour = ifelse(LOCATION == 'Ireland','red',
                                  ifelse(LOCATION == 'OECD - Average','black','#7F7F7F')))

# change subject levels for Ireland and OECD-Average to have a total of 6 subject levels
PISA_2018$SUBJECT[PISA_2018$LOCATION == 'Ireland'] <- c('Boys-IRL','Girls-IRL')
PISA_2018$SUBJECT[PISA_2018$LOCATION == 'OECD - Average'] <- c('Boys-AVG','Girls-AVG')

# convert the SUBJECT and LOCATION columns to factors
PISA_2018$LOCATION <- factor(PISA_2018$LOCATION, levels =  unique(PISA_2018$LOCATION))
PISA_2018$SUBJECT <- factor(PISA_2018$SUBJECT)

```


```{r plot_P2}
# geom_point to plot the Boys and Girls datapoints and speciy the visual attributes
ggplot(PISA_2018) + geom_point(aes(x = LOCATION, y = Value, color = SUBJECT, 
                                   shape = SUBJECT, fill = SUBJECT), size=1.5) + 
  
  # manually specify the shape, fill and color for each of the subject level
  # using breaks in each function to override the legends
  # Levels (in order) : Boys  Boys-AVG    Boys-IRL    Girls   Girls-AVG     Girls-IRL
  scale_shape_manual(values = c(19, 19, 19, 23, 23, 23), breaks = c("Boys", "Girls")) +
  
  scale_fill_manual(values = c("#E2EDF3", "black", "red", "#E2EDF3", "black", "red"),
                    breaks = c("Boys", "Girls")) +
  
  scale_color_manual(values = c("#406D89", "black", "red", "#406D89", "black", "red"),
                     breaks = c("Boys", "Girls")) +

  # pseudo x-axis gridline by joining the points between y=340 and each of the Boys
  geom_segment(aes(x = LOCATION, y = 340, xend = LOCATION, yend = Boys - 3),
               colour = "white", size = 0.2) +
  
  # joining the Boys and Girls data-points with a line segment for each location
  geom_segment(aes(x = LOCATION, y = Boys + 3, xend = LOCATION, yend = Girls - 3),
               colour = "#C4D2DB", size = 0.2) +
  
  # drawing a blue line above the plot as shown in image
  annotate(geom = 'segment', y = Inf, yend = Inf, x = -Inf, xend = Inf, color = 'blue', 
           size = 0.3) +
  
  # specifying axis scales, expanding x-axis to accommodate y-axis text inside the plot 
  scale_x_discrete(expand=c(0.0,2.2)) + 
  scale_y_continuous(breaks = seq(340, 560, 20), limits = c(340, 560)) + 
  
  # adding Title and subtitle
  ggtitle(   " Reading performance (PISA)", 
  subtitle = "Boys / Girls, Mean score, 2018                                                           Source: PISA: Programme for International Student Assessment") +
  
  theme(    # adjusting the axis texts and plot title & subtitles
    axis.text.x = element_text( angle = 45, hjust = 1, size = 6, 
                                     colour = PISA_2018$text_Colour[c(T,F)]),
         
    axis.text.y = element_text(vjust = -0.5,margin = margin(l = 0, r = -20),
                                    size = 6, colour ="#7F7F7F"), 
         
    plot.title = element_text(color = "#615C5C", size = 10, face = "bold", 
                                   family = "Times", hjust=0, margin = margin(b=-6)),
         
    plot.subtitle = element_text(color = "#7F7F7F", size = 6, face = "bold",
                    hjust = 1, vjust = 10, margin = margin(t=10,b=-6),family = "Times"),
    
    # specifying the plot background     
    panel.background = element_rect(fill = "#E2EDF3"),
    
    # adjusting the panel grids   
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(size=0.2),
    
    # adjusting the legend appearance
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.margin = margin(-5,0,0,-380),
    legend.key = element_rect(fill = "white"),
    legend.text = element_text(colour = "#7F7F7F",size = 8),
    
    # removing axis ticks and titles
    axis.ticks = element_blank(),
    axis.title = element_blank())
```
\pagebreak

## Code

```{r eval=FALSE, echo=TRUE, ref.label=c("code_P2","plot_P2")}

```

\pagebreak

# PART 3 : Reading Performance of Countries between 2000-2018

## Explanation

To present the performance of a country I used *Line graph* to show the *trends* of followed by them over the years. To make sure that the data is consistent, I grouped the dataset by *LOCATION & SUBJECT* and calculated the frequency of each. As there are 7 years data, I filtered only those LOCATIONS which had the complete data over the years for both SUBJECTS(Boys & Girls). Once done, I *averaged the Value/PISA scores for each LOCATION* over the 7 years data, and came up with 1 Value for each LOCATION - The average score over the years (OECD-Average is also considered as a LOCATION and is present in the dataset). I arranged them in *descending order to get the ranking of LOCATIONS* and selected top 5 LOCATIONS. 

I plot all target countries with different color and place them in the foreground plot and the remaining LOCATIONS in the background of the plot with a light gray color such that it does not interfere with the target LOCATIONS. I faceted the plot on SUBJECT(Boys and Girls) so that it can be clear how each LOCATION's gender(Boys and Girls) performed over the years and compare them within the top 5 LOCATIONS.

*NOTE : I used the Qualitative Color palettes from DV - Assignment 2 for assigning colors to each of the top 5 Locations.*

## Plot

```{r code_P3}

# group the data by LOCATION & SUBJECT and calculate the frequency. Filter the LOCATIONS
# having complete data over the years.
PISA_comp <- PISA %>% filter(SUBJECT != 'TOT') %>% group_by(LOCATION,SUBJECT)%>%
                 mutate(count = n()) %>% filter(count == 7) %>% ungroup()

# average OECD PISA score for each country over the years
PISA_country_avg <- PISA_comp %>% group_by(LOCATION) %>%
                    summarise(Average = mean(Value)) %>%
                    arrange(desc(Average))

# Top 5 countries which have highest OECD PISA reading average over the years
# finland, Korea, Canada, New Zealand, Ireland
target <- top_n(PISA_country_avg,5)$LOCATION 

# add OECD-Average to the target to help comparison.
target <- c(target, 'OECD - Average')

# subset the data for the target LOCATIONS
PISA_t5_country <- subset(PISA_comp, LOCATION %in% target)

# add color which is assigned to the top 5 LOCATIONS
# unique(PISA_t5_country$LOCATION)
# Canada, Finland, Ireland, Korea, New Zealand, OECD - Average
colour <- c("#8c9900", "#8b99e9", "#b4d025", "#cb3a7d", "#3c3e3c", "red")

```


```{r plot_P3}
#background plot
background <- ggplot() +
  
  #Background lines must be thin and with alpha = 0.6, color matches part 2 theme
  geom_line(PISA_comp, mapping = aes(x = TIME, y = Value, group = LOCATION),
    size = 0.23,
    color = "#C4D2DB",
    alpha = 0.6,
    show.legend = FALSE) +
  
  # facet the plot by SUBJECT
  facet_wrap(~ SUBJECT,scales = "free_y")

# foreground plot
background +
  # Add the lines for the top 5 countries
  geom_line(PISA_t5_country, 
            mapping = aes(x = TIME, y = Value, colour = LOCATION, group = LOCATION, 
                          linetype = LOCATION),
            size = 0.7,alpha = 0.7) +
  
  # Manually assign colors, lineType for each LOCATION
  scale_colour_manual(values = colour) +
  scale_linetype_manual(values = c(rep("solid",5),"dotted")) +
  
  # facet by SUBJECT -> use scale = free_y to add y-axis on both facets
  facet_wrap( ~ SUBJECT,scales = "free_y") + 
  
  # add blue line above the plot - same as part 2
  annotate(geom = 'segment', y = Inf, yend = Inf, x = -Inf, xend = Inf, color = 'blue', 
           size = 0.3) +
  
  # specifying axis scales - same as plot 1 
  scale_y_continuous( breaks = seq(340, 600, 20), limits = c(340, 600)) + 
  scale_x_continuous( breaks = seq(2000, 2018, 3),
                      labels = seq(2000, 2018, 3),
                      expand = c(0, 0.1)) +
  
  # add plot title and subtitle 
  ggtitle(   " Reading performance (PISA) from 2000 - 2018", 
  subtitle = "  Highlighting performance of Top 5 Countries                                                            Source: PISA: Programme for International Student Assessment") + 
  
  #Using same theme as part 2
  theme( # adjusting the axis text, plot title and subtitle appearance 
    axis.text = element_text( size = 5.5, colour = "#7F7F7F"),
    
    plot.title = element_text(color = "#615C5C", size = 11, face = "bold", 
                              family = "Times", hjust=0),
         
    plot.subtitle = element_text(color = "#7F7F7F", size = 7, face = "bold",
                                 family = "Times"),
    
    # specifying plot background
    panel.background = element_rect(fill = "#E2EDF3"),

    # adjusting panel grids
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(size=0.2),
       
    # adjusting facets text and background fill
    strip.background = element_rect(fill = "#5f89a3"),
    strip.text.x = element_text(face = "bold", size = 8),
         
    # adjusting the legend appearance     
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.key = element_rect(fill = "white"),
    legend.text = element_text(color = "#7F7F7F",size = 8),
    
    # removing axis ticks and titles
    axis.ticks = element_blank(),
    axis.title = element_blank()) + 
  
  # adjusting the legend to make it appear in 1 single row
  guides(colour = guide_legend(nrow = 1))
```
\pagebreak

## Code

```{r eval=FALSE, echo=TRUE, ref.label=c("code_P3","plot_P3")}

```

# References

- Lecture slides
- GGPLOT2 Notes
- https://stackoverflow.com/questions/2375587/reorder-levels-of-a-factor-without-changing-order-of-values
- https://stackoverflow.com/questions/19506630/control-point-border-thickness-in-ggplot
- https://stackoverflow.com/questions/58380045/plotting-lines-between-two-points-in-ggplot2
- https://stackoverflow.com/questions/13461829/select-every-other-element-from-a-vector
- http://search.r-project.org/library/ggplot2/html/expansion.html
- https://stackoverflow.com/questions/55406829/ggplot-put-axis-text-inside-plot
- https://www.datanovia.com/en/blog/ggplot-title-subtitle-and-caption/
- https://stackoverflow.com/questions/46256851/how-to-add-line-at-top-panel-border-of-ggplot2
- https://stackoverflow.com/questions/30076911/how-do-i-display-only-selected-items-in-a-ggplot2-legend
- https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
- https://stackoverflow.com/questions/36087262/ggplot2-legend-items-in-a-single-horizontal-row

**NOTE : The colors used were selected by an firefox extension: colorPick Eyedropper. - This extension picks the pixel colour from an image and returns HEX value to be used in plots**